Acceleration of Tropical Cyclones As a Proxy For Extratropical Interactions: Synoptic Scale Patterns and Long-Term Trends

### 2020-11-12 paper/manuscript.tex
update

\introduction
The motion of a tropical cyclone (TC), to a leading order, is known to be  governed by the large-scale environment surrounding it \citep[e.g.][]{Hodanish1993}. It was recognized early on that the forward speed of a TC can be closely approximated by the surrounding wind field \citep{E2018}.  It is also understood that a TC is not an isolated vortex that is passively carried by a current. The background environment also has attendant gradients of potential vorticity, moisture, and wind shear. A TC actively responds to these external stimuli. Consequently, this impacts not only the motion and intensity of the TC, but also the the large scale flow. For instance, the generation of the $\beta-gyres$, which infuences TC motion, is a response of the TC circulation to the background potential vorticity gradient. Similarly, vertical wind shear can modulate the intensity of the TC via ventilation of low entropy air into the storm core. This, in turn,  can also impact subsequent TC motion. On the other hand, a TC can influence the large-scale flow by exciting  waves in the extratropical stormtrack, leading to rapid downstream development and rearrangement of the flow. Thus, it can be contended that a TC is always interacting with its environment and the interaction is, in part, manifested in its motion.

@@

To calculate the curvature acceleration, it is necessary to first determine the radius of curvature, R. A standard approach to calculating R, given a set of discrete points along a curve (in our case, a TC track), is to fit a circle through three consecutive points. For a curved line on a sphere, it can be shown that:

@@

The corresponding information for western North Pacific for 1732 storms that met the 2-day threshold is presented in the appendix (Figure \ref{fig:distWP} and Table \ref{tab:distWP}). In this case, we consider all track points, including NR and ET for reasons discussed in section 2. The basic distribution of speed and accelerations are similar to the North Atlantic except that the storms, on average, were found to have negative tangential acceleration poleward of 40$^\circ$N.

%In essence, we draw a distinction between the propagation of extratropical and tropical systems. We reason that an extratropical cyclone embedded in the stormtrack is subject to baroclinic development and dynamics of Rossby wavepackets \citep[e.g.][]{K2019}.  In contrast, a tropical system is primarily subject to steering by the background flow. \cite The importance of this distinction will be emphasized further in a subsequent section.

@@

Figure \ref{fig:tang_cdf} compares the cumulative probability distribution (CDF) of tangential accelerations over two 10-year periods: 1966--1975 and 2010--2019. Three latitude bands are shown - the entire Atlantic,  0--30$^o$N, and 20--40$^o$N. The CDFs show that the median value of tangential accelertion does not change much between the two decades. Notably, however, there is a clear reduction in the frequency of rapid acceleration and deceleration. Figure \ref{fig:curv_cdf} shows CDFs for curvature acceleration. Here too, there is a similar shift towards less frequent rapid acceleration in the past decade compared to 1966--1975. There is a small compensating increase in frequency of relatively less rapid acceleration values. In the case of  0--20$^o$N, the increase is seen for values below the 90$^\text{th}$\% while for 20--40$^o$N, it is seen for values below the median.

In the preceding sections, we showed that rapid acceleration or deceleration of a TC is typically associated with an interaction with extratropical stormtrack. It is of interest to see if the difference in the CDF is related to a long-term trend in TC accleration. The motivation being that it can inform us regarding potential trends in the nature of TC-baroclinic stormtrack interaction. For this, we consider rapid acceleration and deceleration separately. We do this for two reasons. First, as seen in Fig.  \ref{fig:tang_cdf}, the median acceleration does not change significantly and this is also the case for the mean acceleration (not shown). Second, the synoptic patterns associated with rapid acceleration and rapid deceleration of TCs are entirely different. We consider three threshold values to calculate trends. These are 90$^\text{th}$\%, 80$^\text{th}$\% and 60$^\text{th}$\% values of both categories. The trends over 1966--2019 are calculated using the annual-average acceleration exceeding these threhsolds for for Atlantic TCs after excluding instances of NR and ET. The trends and the associated P-value calculated using two different methods - (a) linear regression, and (b) the Thiel-Sen linear estimate with the Mann-Kendal significance.

@@

@@

\clearpage
%========================================================
\begin{figure*}[t]
  \centering
    \includegraphics[width=.85\textwidth]{composite_tang_40_30_multilevel_NA.png}
  \caption{Storm-relative composite average geopotential heights (thick orange lines) and anomalies (color shaded) for all TCs located in the latitude bin 30-40$^o$N over the Atlantic. The composite fields are shown for three levels -- 300 hPa, 500 hPa and 850 hPa. In each panel, the composite 1000 hPa anomalous geopotential is shown using thin black contours. All anomalies are defined relative to a long-term synoptic climatology. The contour intervals are: 12 dam, 6 dam and 3 dam for the three levels respectively. The shading interval in dam for the 300 hPa anomaly fields is shown in the label-bar. It is half of the value for the other two levels. The left column is for rapid tangential acceleration and the right column is for rapid tangential deceleration.}
  \label{fig:tangComp}
\end{figure*}
%========================================================

@@

\begin{figure*}[t]
  \includegraphics[width=14cm]{ET_relative_accel_speed.png}
  \caption{Composite speed and accelerations relative to time of ET. A single pass of 5-point running average was applied to the speed and tangential acceleration curves. Two passes of the same filter were applied to the curvature acceleration.}\label{fig:ET_SA}
\end{figure*}
%========================================================

@@

\begin{figure*}[t]
  \includegraphics[width=6cm]{cdf_tang_atlWide.png}
  \includegraphics[width=6cm]{cdf_tang_eq_20N.png}
  \includegraphics[width=6cm]{cdf_tang_20_40N.png}

  \caption{CDF}\label{fig:tang_cdf}
\end{figure*}

\begin{figure*}[t]
  \includegraphics[width=6cm]{cdf_curv_atlWide.png}
  \includegraphics[width=6cm]{cdf_curv_eq_20N.png}
  \includegraphics[width=6cm]{cdf_curv_20_40N.png}

  \caption{CDF}\label{fig:curv_cdf}
\end{figure*}
%========================================================

>>>>

\introduction

The motion of a tropical cyclone (TC) is known to be  governed by the large-scale environment surrounding it \citep[e.g.][]{Hodanish1993}. It was recognized early on that the forward speed of a TC can be closely approximated by the surrounding wind field \citep{E2018}.  It is also understood that a TC is not an isolated vortex that is passively carried by a current. The background environment also has attendant gradients of potential vorticity, moisture, and wind shear. A TC actively responds to these external stimuli. Consequently, this impacts not only the motion and intensity of the TC, but also the the large scale flow. For instance, the generation of the $\beta$-\emph{gyres} -- that influence TC motion -- is a response to the background potential vorticity gradient (citation). Similarly, vertical wind shear can modulate the intensity of the TC via ventilation of low entropy air into the storm core (citation) . This, in turn,  can also impact subsequent TC motion. On the other hand, a TC can influence the large-scale flow by exciting  waves in the extratropical stormtrack, leading to rapid downstream development and rearrangement of the flow. Thus, it can be contended that a TC is always interacting with its environment and the interaction is, in part, manifested in its motion.

@@

To calculate the curvature acceleration, it is necessary to first determine the radius of curvature, R. A standard approach to calculating R, given a set of discrete points along a curve -- in our case, a TC track -- is to fit a circle through three consecutive points. For a curved line on a sphere, it can be shown that:

@@

Figure \ref{fig:distNAMonth} shows the distribution of mean tangential and curvature accelerations within the latitude band 20--40$^oN$ as a function of month. The numbers on top of the bars show the percentage of the total number of track points over all months. As expected, September shows the highest percentage of track points, followed by August and October. During the peak Atlantic hurricane season, the magnitude of the mean tangential and curvature accelerations increase steadily from July--November.

%The corresponding information for western North Pacific for 1732 storms that met the 2-day threshold is presented in the appendix (Figure \ref{fig:distWP} and Table \ref{tab:distWP}). In this case, we consider all track points, including NR and ET for reasons discussed in section 2. The basic distribution of speed and accelerations are similar to the North Atlantic except that the storms, on average, were found to have negative tangential acceleration poleward of 40$^\circ$N. %In essence, we draw a distinction between the propagation of extratropical and tropical systems. We reason that an extratropical cyclone embedded in the stormtrack is subject to baroclinic development and dynamics of Rossby wavepackets \citep[e.g.][]{K2019}.  In contrast, a tropical system is primarily subject to steering by the background flow. \cite The importance of this distinction will be emphasized further in a subsequent section.

@@

Figure \ref{fig:tang_cdf} compares the cumulative probability distribution (CDF) of tangential accelerations taken from all months over two 10-year periods: 1966--1975 and 2010--2019. Three regions are shown: Entire Atlantic,  0--30$^o$N, and 20--40$^o$N. The CDFs show that the median value of tangential acceleration does not change much between the three decades. Notably, however, there is a clear reduction in the frequency of rapid acceleration and deceleration. Figure \ref{fig:curv_cdf} shows CDFs for curvature acceleration. Here too, there is a similar shift towards less frequent rapid acceleration in the past decade compared to 1966--1975. There is a small compensating increase in frequency of relatively less rapid acceleration values. In the case of  0--20$^o$N, the increase is seen for values below the 90$^\text{th}$\% while for 20--40$^o$N, it is seen for values below the median.

As noted earlier from Fig. \ref{fig:distNAMonth}, the magnitude of accelerations increases steadily going from July--November. Fig

CDFs for individual months to get a sense of how the

In the preceding sections, we showed that rapid acceleration or deceleration of a TC is typically associated with an interaction with extratropical stormtrack. The synoptic-scale pattern associated with these two extremes are distinct in the phasing of the TC and the extratropical wavepacket. It is of interest to see if the differences of the CDFs (Figure \ref{fig:tang_cdf}) are related to long-term trends in TC acceleration. The motivation being that it can inform us about potential trends in the nature of TC-baroclinic stormtrack interaction. For this, we consider rapid acceleration and deceleration separately. We do this for two reasons. First, as seen in Fig.  \ref{fig:tang_cdf}, the median acceleration does not change significantly and this is also the case for the mean acceleration (not shown). Second, the synoptic patterns associated with rapid acceleration and rapid deceleration of TCs are entirely different. We consider three threshold values to calculate trends. These are 90$^\text{th}$\%, 80$^\text{th}$\% and 60$^\text{th}$\% values of both categories. The trends over 1966--2019 are calculated using the annual-average acceleration exceeding these threhsolds for for Atlantic TCs after excluding instances of NR and ET. The trends and the associated P-value calculated using two different methods - (a) linear regression, and (b) the Thiel-Sen linear estimate with the Mann-Kendal significance.

@@

account for:

1) During the peak Atlantic hurricane season, the magnitude of the mean tangential and curvature accelerations increase steadily from July--November.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=.45\textwidth]{tang_mon_climo.png}
  \includegraphics[width=.45\textwidth]{curv_mon_climo.png}
  \caption{Distribution of tangential and curvature accelerations of Atlantic TCs within the latitude band 20--40$^\circ$N
    as a function of month. Numbers on top of the bars show the percentage of track points that occurred during each month.
    Storm instances classified as ET or NR were excluded}
  \label{fig:distNAMonth}
\end{figure*}

\begin{figure*}[t]
  \centering
    \includegraphics[width=.85\textwidth]{composite_tang_40_30_multilevel_NA.png}
  \caption{Storm-relative composite average geopotential heights (thick orange lines) and anomalies (color shaded) for all TCs located in the latitude bin 30-40$^o$N over the Atlantic. The composite fields are shown for three levels -- 300 hPa, 500 hPa and 850 hPa. In each panel, the composite 1000 hPa anomalous geopotential is shown using thin black contours. All anomalies are defined relative to a long-term synoptic climatology. The contour intervals are: 12 dam, 6 dam and 3 dam for the three levels respectively. The shading interval in dam for the 300 hPa anomaly fields is shown in the label-bar. It is half of the value for the other two levels. The left column is for rapid tangential acceleration and the right column is for rapid tangential deceleration.}
  \label{fig:tangComp}
\end{figure*}
%========================================================

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ET_relative_accel_speed.png}
  \caption{Composite speed and accelerations relative to time of ET. A single pass of 5-point running average was applied to the speed and tangential acceleration curves. Two passes of the same filter were applied to the curvature acceleration.}\label{fig:ET_SA}
\end{figure*}
%========================================================

@@

%========================================================
%Figure 
%========================================================
\begin{figure*}[t]
  \includegraphics[width=14cm]{tangential_cdf.png}
  \caption{CDF}\label{fig:tang_cdf}
\end{figure*}
%========================================================

%========================================================
%Figure 
%========================================================
\begin{figure*}[t]
  \includegraphics[width=14cm]{aug_cdf_atl.png}
  \includegraphics[width=14cm]{sep_cdf_atl.png}
  \includegraphics[width=14cm]{oct_cdf_atl.png}
  \caption{CDF}\label{fig:tang_month_cdf}
\end{figure*}
%========================================================

%Figure 
%========================================================

\begin{figure*}[t]
  \includegraphics[width=14cm]{curv_cdf.png}
  \caption{CDF}\label{fig:curv_cdf}
\end{figure*}
%========================================================
